### 2023-03-02 calculus.tex
Edit text in relation to type error mosaic figure

Programmers often work with program states that are syntactically well-formed but not well-typed.
\Cref{fig:calculus-examples} shows common type errors as they appear in the Hazel programming
environment. For example, a simple error that may occur is the use of an unbound variable. As in
\cref{fig:calculus-examples-unbound}, editor services indicate the usage of unbound $y$ is invalid
by highlighting it in red.

@@

Sometimes, the type an expression \emph{should} be is known. Consider
\cref{fig:calculus-examples-inconsistent-types}, in which it is expected that both operands of the
$+$ operator are numbers. Since $\textsf{true}$ is not a number, a type error arises, and Hazel
indicates a mismatch between the expected and actual types of the operand. The application in
\cref{fig:calculus-examples-app-non-lambda} of a non-lambda leads to a similar kind of error. In
this case, however, the expectation is not a singular type, but any member of a family of function
types. \Cref{fig:calculus-examples-inconsistent-branches} presents an error of inconsistent
branches: the branches of the if-else expression are of different types. Hazel opts to highlight the
entire expression in such a situation, while other tools may choose one branch to be ``correct'' and
``blame'' the type mismatch error on the other.

@@

\begin{figure}[htbp]
  \[\begin{array}{rrcl}
    \TMName  & \TMV  & \Coloneqq & \TUnknown \mid \TNum \mid \TBool \mid \TArrow{\TMV}{\TMV} \mid \TProd{\TMV}{\TMV} \\
    \EMName  & \EMV  & \Coloneqq & x \mid \ELam{x}{\TMV}{\EMV} \mid \EAp{\EMV}{\EMV} \mid \ELet{x}{\EMV}{\EMV}
                       \mid           \ENumMV \mid \EPlus{\EMV}{\EMV} \\
             &       & \mid         & \ETrue \mid \EFalse \mid \EIf{\EMV}{\EMV}{\EMV}
                       \mid           \EPair{\EMV}{\EMV}
                       \mid           \EProjL{\EMV} \mid \EProjR{\EMV}
                       \mid           \EEHole \\
    \ECMName & \ECMV & \Coloneqq & x \mid \ECLam{x}{\TMV}{\ECMV} \mid \ECAp{\ECMV}{\ECMV} \mid \ECLet{x}{\ECMV}{\ECMV}
                       \mid           \ECNumMV \mid \ECPlus{\ECMV}{\ECMV} \\
             &       & \mid         & \ECTrue \mid \ECFalse \mid \ECIf{\ECMV}{\ECMV}{\ECMV}
                       \mid           \ECPair{\ECMV}{\ECMV} \mid \ECProjL{\ECMV} \mid \ECProjR{\ECMV}
                       \mid           \ECEHole \\
             &       & \mid         & \ECUnbound{x} \mid \ECInconType{\ECMV} \mid \ECInconBr{\ECMV}{\ECMV}{\ECMV} \mid \ECInconAsc{\ECMV} \\
             &       & \mid         & \ECSynNonMatchedArrow{\ECMV} \mid \ECAnaNonMatchedArrow{\ECMV}
                       \mid           \ECSynNonMatchedProd{\ECMV} \mid \ECAnaNonMatchedProd{\ECMV}
  \end{array}\]
  \caption{Syntax of an extended GTLC.}
  \label{fig:calculus-syntax}
\end{figure}

@@

Again, to satisfy totality, we must consider the case when $\TMV$ is not a matched arrow type, such
as in the example of \cref{fig:calculus-examples-app-non-lambda}. There is no expected type for the
argument, so we perform analytic marking on $\EMV_2$ against the unknown type. In this case, it is
not quite right to mark $\EMV_1$ with the inconsistent type mark; rather than any single type, it is
a member of the family of function types that is expected. For such a ``constrained'' synthetic
mode, we use a specialized mark $\ECSynNonMatchedArrow{\ECMV}$, which indicates that $\ECMV$ was
expected to be a function---but was not. The output type is unknown, so the entire expression
synthesizes the unknown type.
%
\begin{mathpar}
  \judgment{
    \ctxSynFixedInto{\ctx}{\EMV_1}{\ECMV_1}{\TMV} \\
    \notMatchedArrow{\TMV} \\
    \ctxAnaFixedInto{\ctx}{\EMV_2}{\ECMV_2}{\TUnknown}
  }{
    \ctxSynFixedInto{\ctx}{\EAp{\EMV_1}{\EMV_2}}{\ECApSynNonMatchedArrow{\ECMV_1}{\ECMV_2}}{\TUnknown}
  }{MKSAp2}

@@

The framework also allows for other localization approaches. For example, we may regard the first
branch as ``correct'' and localize any errors to the second. The following rule formalizes such a
design:
\[%
  \judgment{
    \ctxAnaFixedInto{\ctx}{\EMV_1}{\ECMV_1}{\TBool} \\
    \ctxSynFixedInto{\ctx}{\EMV_2}{\ECMV_2}{\TMV} \\
    \ctxAnaFixedInto{\ctx}{\EMV_3}{\ECMV_3}{\TMV}
  }{
    \ctxSynFixedInto{\ctx}{\EIf{\EMV_1}{\EMV_2}{\EMV_3}}{\ECIf{\ECMV_1}{\ECMV_2}{\ECMV_3}}{\TMV}
  }{MKSIf'}
\]%
Of course, corresponding unmarked and marked typing rules would follow similar structure, and other
heuristics may be devised for different kinds of branching structures.

>>>>

Programmers often work with program states that are syntactically well-formed but not well-typed.
\Cref{fig:calculus-examples} shows common type errors as they appear in the Hazel programming
environment. For example, a simple error that may occur is the use of an unbound variable. As in
\cref{fig:calculus-examples-unbound}, the editor indicates that the usage of an unbound $y$ is
invalid by highlighting it in red.

@@

Sometimes, the type an expression \emph{should} be is known. Consider
\cref{fig:calculus-examples-inconsistent-types}, in which it is expected that both operands of the
$+$ operator are numbers. Since $\textsf{true}$ is not a number, a type error arises, and Hazel
indicates a mismatch between the expected and actual types of the operand. The application in
\cref{fig:calculus-examples-app-non-lambda} of a non-lambda leads to a similar kind of error. In
this case, however, the expectation is not a singular type, but any member of a family of function
types. \Cref{fig:calculus-examples-inconsistent-branches} presents an error of inconsistent
branches: the branches of the conditional are of different types: $5$ is a number, while
$\textsf{false}$ is a boolean. Hazel opts to highlight the entire expression in such a situation,
while other tools may choose one branch to be ``correct'' and ``blame'' the type mismatch error on
the other.

@@

\begin{figure}[htbp]
  \[\begin{array}{rrcl}
    \TMName  & \TMV  & \Coloneqq & \TUnknown \mid \TNum \mid \TBool \mid \TArrow{\TMV}{\TMV} \mid \TProd{\TMV}{\TMV} \\
    \EMName  & \EMV  & \Coloneqq & x \mid \ELam{x}{\TMV}{\EMV} \mid \EAp{\EMV}{\EMV} \mid \ELet{x}{\EMV}{\EMV}
                       \mid           \ENumMV \mid \EPlus{\EMV}{\EMV} \\
             &       & \mid         & \ETrue \mid \EFalse \mid \EIf{\EMV}{\EMV}{\EMV}
                       \mid           \EPair{\EMV}{\EMV}
                       \mid           \EProjL{\EMV} \mid \EProjR{\EMV}
                       \mid           \EEHole \\
    \ECMName & \ECMV & \Coloneqq & x \mid \ECLam{x}{\TMV}{\ECMV} \mid \ECAp{\ECMV}{\ECMV} \mid \ECLet{x}{\ECMV}{\ECMV}
                       \mid           \ECNumMV \mid \ECPlus{\ECMV}{\ECMV} \\
             &       & \mid         & \ECTrue \mid \ECFalse \mid \ECIf{\ECMV}{\ECMV}{\ECMV}
                       \mid           \ECPair{\ECMV}{\ECMV} \mid \ECProjL{\ECMV} \mid \ECProjR{\ECMV}
                       \mid           \ECEHole \\
             &       & \mid         & \ECUnbound{x} \mid \ECInconType{\ECMV} \mid \ECInconBr{\ECMV}{\ECMV}{\ECMV} \mid \ECInconAsc{\ECMV} \\
             &       & \mid         & \ECSynNonMatchedArrow{\ECMV} \mid \ECAnaNonMatchedArrow{\ECMV}
                       \mid           \ECSynNonMatchedProd{\ECMV} \mid \ECAnaNonMatchedProd{\ECMV}
  \end{array}\]
  \caption{Syntax of an extended GTLC and its marked version.}
  \label{fig:calculus-syntax}
\end{figure}

@@

Again, to satisfy totality, we must consider the case when $\TMV$ is not a matched arrow type, such
as in the example of \cref{fig:calculus-examples-app-non-lambda}. There is no expected type for the
argument, so we perform analytic marking on $\EMV_2$ against the unknown type. In this case, it is
not quite right to mark $\EMV_1$ with the inconsistent type mark; rather than any single type, it is
any type within the family of arrow types that is expected. For such a ``constrained'' synthetic
mode, we use a specialized mark $\ECSynNonMatchedArrow{\ECMV}$, which indicates that $\ECMV$ was
expected to be a function---but was not. The output type is unknown, so the entire expression
synthesizes the unknown type.
%
\begin{mathpar}
  \judgment{
    \ctxSynFixedInto{\ctx}{\EMV_1}{\ECMV_1}{\TMV} \\
    \notMatchedArrow{\TMV} \\
    \ctxAnaFixedInto{\ctx}{\EMV_2}{\ECMV_2}{\TUnknown}
  }{
    \ctxSynFixedInto{\ctx}{\EAp{\EMV_1}{\EMV_2}}{\ECApSynNonMatchedArrow{\ECMV_1}{\ECMV_2}}{\TUnknown}
  }{MKSAp2}

@@

The framework also allows for other localization approaches. For example, as discussed at the
beginning of the section, we may regard the first branch as ``correct'' and localize any errors to
the second. The following rule formalizes such a design:
\[%
  \judgment{
    \ctxAnaFixedInto{\ctx}{\EMV_1}{\ECMV_1}{\TBool} \\
    \ctxSynFixedInto{\ctx}{\EMV_2}{\ECMV_2}{\TMV} \\
    \ctxAnaFixedInto{\ctx}{\EMV_3}{\ECMV_3}{\TMV}
  }{
    \ctxSynFixedInto{\ctx}{\EIf{\EMV_1}{\EMV_2}{\EMV_3}}{\ECIf{\ECMV_1}{\ECMV_2}{\ECMV_3}}{\TMV}
  }{MKSIf'}
\]%
Of course, corresponding unmarked and marked typing rules would follow similar structure, and other
heuristics may be devised for different kinds of branching structures.
